Sequential rank agreement methods for comparison of lists

### 2015-08-05 latex/paper.Rnw
Application afsnit opdateret

\section{Applications}

We now consider an application of the sequential rank agreement to two
data sets consisting of MALDI-TOF (Matrix-Assisted Laser
Desorption/Ionization Time Of Flight) mass spectra obtained from blood
samples from patients with either benign or malignant ovarian
tumors. The data sets are subsamples of the Danish MALOVA and DACOVA
study populations.

The MALOVA study is a multidisciplinary Danish study on ovarian cancer
\citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999. For the purpose of illustration we use a random
subsample of $119$ patients with a total of $58$ patients with
malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another
multidisciplinary Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\citep{bertelsen1991protocol}. It was aimed to continuously enroll all
patients that were referred to surgery of an ovarian tumor clinically
suspected to be cancer during the period 1984 to 1990.
% @Andreas: hvorfor random subsample?
Similarly, we use a random subsample of $113$
patients with a total of $54$ malignant ovarian cancers and $59$
benign ovarian tumors/gynecologic disorders.

Proteomic analysis was performed by MALDI-TOF mass spectrometry with a
Bruker Ultraflex in positive, linear ion mode in the mass-to-charge
range $800$ to $20000$ Dalton by an average of $500$ laser
shots. Calibration of the instrument was performed using a protein
standard mixture. Each spectrum consists of $49642$ samples over the
mass-to-charge range which we down-sample on an equidistant grid of
5000 points by linear interpolation. We further preprocess each
spectrum by first removing the slow-varying baseline intensities with
the SNIP algorithm \citep{ryan1988snip} followed by a normalization
with respect to the total ion counts. Finally, we standardize the 5000
predictors to have column-wise zero mean and unit variance.

We use these two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of different statistical
classification methods in how they rank predictors by different
statistical classification methods for the two study populations. In
the second scenario we assess the agreement among ranking of the
predicted individual risk probabilities of having malignant tumor
using four different prediction methods.
For this second example we furthermore
investigate the behaviour of different statistical methods in a
simulation study.

\subsection{Agreement of predictor ranking}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/marginalVsModel.pdf}
\end{center}
 \caption{Marginal}
 \label{fig:app1}
\end{figure}

\subsection{Agreement of predictions}

Here we apply the same statistical classification
method to XX bootstrap samples of the MALOVA data to obtain four
prediction models. From each model we predict the risks of a malignant
patients in the DACOVA study.  Finally we estimate the agreement of
the resulting XX rankings of the DACOVA patients to illustrate the
sensitivity of the corresponding statistical classification method to
pertubations of the learning data.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/thomas.pdf}
\end{center}
 \caption{Thomas}
 \label{fig:app2}
\end{figure}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/simulationTest2.pdf}
\end{center}
 \caption{SimulationTest2}
 \label{fig:app3}
\end{figure}

>>>>

\section{Application}
We now consider an application of the sequential rank agreement to two data sets consisting of MALDI-TOF (Matrix-Assisted Laser Desorption/Ionization Time Of Flight) mass spectra obtained from blood samples from patients with either benign or malignant ovarian tumors. The data sets are subsamples of the Danish MALOVA and DACOVA study populations.

The MALOVA study is a multidisciplinary Danish study on ovarian cancer \citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed with an ovarian tumor and referred for surgery from the participating departments of gynecology were enrolled continuously from December 1994 to May 1999. For the purpose of illustration we use a random subsample of $119$ patients with a total of $58$ patients with malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another multidisciplinary Danish study on ovarian cancer which included about $66\%$ of the female population of Denmark \citep{bertelsen1991protocol}. It was aimed to continuously enroll all patients that were referred to surgery of an ovarian tumor clinically suspected to be cancer during the period 1984 to 1990. Similarly, we use a random subsample of $113$ patients with a total of $54$ malignant ovarian cancers and $59$ benign ovarian tumors/gynecologic disorders.

Each spectrum consists of $49642$ samples over a mass-to-charge range of $800$ to $20000$ which we down-sample on an equidistant grid of 5000 points by linear interpolation. We preprocess each spectrum by first removing the slow-varying baseline intensity with the SNIP algorithm \citep{ryan1988snip} followed by a normalization with respect to the total ion count. We then standardize the 5000 predictors to have column-wise zero mean and unit variance in each data set.

We use the two data sets to illustrate how the sequential rank agreement can be applied in two different scenarios. In the first scenario we assess the agreement of four different statistical classification methods in how they rank the predictors according to their importance with respect to the binary outcome. In the second scenario we assess the agreement among rankings of individual predicted risk probabilities of having a malignant tumor using the four methods. The first scenario is relevant in the context of biomarker discovery and the latter is important when e.g., ranking patients according to immediacy of treatment.

We use the following four classification methods: Random Forest \citep{breiman2001random} implemented in the R package \texttt{randomForest} \citep{liaw2002classification}, logistic Lasso \citep{tibshirani1996regression} and Ridge regression \citep{segerstedt1992ordinary} both implemented in the R package \texttt{glmnet} \citep{Friedman:2010aa}, and Partial Least Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls} implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

In both scenarios we use the MALOVA data to train the statistical models, and the two types of agreements are assessed with respect to perturbations of the training data in the following manner. We repeatedly draw a sub-sample (without replication) consisting of 90\% of the MALOVA observations and train the four models on the sub-sample. Each of these models produces a ranking of the 5000 predictors according to their importance with respect to discriminating between  the tumor types, and the agreement of these rankings across the different sub-samples are used to assess the stability of the importance of the predictors. For the Random Forest classifier the predictors are ranked according to the Gini index, while for the logistic Lasso and Ridge regression models we order the predictors by absolute magnitude of the estimated regression coefficients. For the PLS-DA model the importance of the predictors is based on a weighted sum of the absolute coefficients where the weights are proportional to the reduction in the sums of squares across the components. To assess the stability of the individual risk predictions we apply the predictors from the DACOVA data set to each of the models. These predicted probabilities are then ranked in decreasing order, and the agreement of the risk predictions are similarly obtained by the sequential rank agreement across the different sub-samples of the training data. We use 1000 iterations for the sub-sampling procedure.

All four models depend on tuning parameters. We estimate these by a 20-times repeated 5-fold internal cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

\begin{figure}[tbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
 \caption{Sequential rank agreement of predicted risk probabilities for the DACOVA data (left) and estimated predictor importance (right) based on 1000 sub-samples of the MALOVA training data.}
 \label{fig:app1}
\end{figure}

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note here that the predictor importance estimated with the logistic Lasso regression model is capable of producing censored lists depending on the value of the penalty parameter. With the deviance criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curve for the importance of the predictors.

For most of the sequential rank agreement curves we see as expected that they start low followed by an increase until they approximately become constant.  This has the interpretation that the classification methods show higher agreement with respect to the highest ranked items (the most high-risk patients and the most important predictors respectively) then for the items deeper in the lists across the different sub-samples. The change-point where the curves become approximately constant is the list depth where the ranks of the remaining items become almost random. The only exceptions are the predictor importance agreements using the logistic Ridge regression models and the risk probability agreement using the logistic Lasso regression with the AUC criterion which show higher disagreement in the top of the lists. The reason for this in the Lasso case is that the AUC criterion tends to favor a very high value of the penalty parameter causing only a single predictor to be selected in all of the 1000 iterations. This results in a lack of generalizability to the DACOVA data  giving rise to the higher disagreement in the predicted risk probabilities. In the extreme case where the penalty becomes so high that none of the predictors are selected by the Lasso, the sequential rank agreement for the predicted probabilities becomes undefined since all the ranks will be ties.

From Figure \ref{fig:app1} it can also be seen that some of the methods appear to be better at agreeing with respect to the predicted probabilities than at agreeing with respect to the ranking of the importance of the predictors and vice versa. Ridge regression performs very well for the risk predictions but is outperformed by PLS-DA for the predictor importances. On the other hand, PLS-DA appears to be similarly worse for the risk predictions compared to Ridge regression. The Random Forest method seems to perform equally well in both scenarios. Lasso regression (except with the AUC criterion as discussed previously) performs very well for the high-risk patients but very bad for the importance of the predictors. This reason for this is the high auto-correlation between the intensities in the mass spectra which leads to collinearity in the regression model. It is well-known that variable selection with the Lasso penalty does not preform very well when the predictors are highly correlated. The collinearity does, however, not affect the risk predictions, since it is not so important what specific variable that get selected from a group of highly correlated predictors when the purpose is risk predictions.

Finally, it would appear that Ridge regression with the AUC criterion for the penalty parameter achieves the best performance with respect to the stability of ranking the individual predicted risk probabilities. It must, however, be stressed that sequential rank agreement is only concerned with whether similar orderings of the patients is obtained under perturbations of the training data. It is not concerned with whether such an ordering is the correct one. This is similarly to the usual distinction between accuracy and precision. Looking at the distributions of the AUCs across the $1000$ iterations for each of the models with respect to the known class labels of the DACOVA data we see that the PLS-DA model attained the best AUC with a median value of $0.71$ (IQR = 0.07) while the Ridge regression with AUC criterion has a median AUC of $0.49$ (IQR = 0.02). This means that while Ridge regression with the AUC criterion achieves the highest rank agreement it performs similar to a random coin toss with respect to classification. Both concerns are of course important in practice and for this data set it appears that PLS-DA is the best method with respect to both concerns.
